Shift-sensitive Power Transformations for Transforming Data to Normality

### 2023-03-02 manuscript/manuscript.Rmd
Updated robust optimisation comparison.

.plot_weighting_functions(plot_theme = plot_theme)
```

@@

The result is shown in Figure **INSERT**, and summarised in Table **INSERT**. For each distribution, transformation parameter $\hat{\lambda}$ varies with $d$ when estimated using the original definitions. In contrast, estimation of $\lambda$ using shift-sensitive power transformations was invariant to $d$.

```{r, results="asis"}
table_data <- data.table::dcast(
  aggregate_data,
  distribution ~ version + method,
  value.var = "text",
  sep = " ")

data.table::setnames(
  table_data,
  old = c("original Box-Cox", "original Yeo-Johnson", "shift-sensitive Box-Cox", "shift-sensitive Yeo-Johnson"),
  new = c("orig. Box-Cox", "orig. Yeo-Johnson", "shift-sens. Box-Cox", "shift-sens. Yeo-Johnson"))

knitr::kable(
  table_data,
  align = c("l", "c", "c", "c", "c"),
  caption = paste0(
    "Estimates and variability of transformation parameter $\\lambda$ under location shifts. ",
    "Whereas the original (orig.) power transformations show considerable variability, ",
    "whereas for shift-sensitive (shift-sens.) power transformations variability in the value of $\\lambda$ is negligible."),
  label = "shifted-distributions-table",
  format = "latex",
  escape = FALSE,
  booktabs = TRUE)
```

@@

```{r optimal_criteria, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Things."
)

# This find the optimal weighting parameter settings that are used within the
# package.
two_sided_function_parameters <- .get_optimised_weighting_function_parameters_two_sided(manuscript_dir = manuscript_dir)

# This finds lambda values at the optimised weighting parameter settings.
data <- .get_transformation_parameters_two_sided(manuscript_dir = manuscript_dir)

data <- data[estimation_method == "MLE"]
data <- data[, "lambda_error" := abs(lambda - target_lambda)]
non_robust_data <- data[weight_method == "non-robust", mget(c("method", "estimation_method", "k", "ii", "lambda_error"))]
data[non_robust_data, on=c("method", "estimation_method", "k", "ii"), "non_robust_lambda_error" := i.lambda_error]

p <- ggplot2::ggplot(
  data = data,
  mapping = ggplot2::aes(
    x = weight_method,
    y = lambda_error,
    fill = method))
p <- p + plot_theme
p <- p + ggplot2::geom_violin(draw_quantiles = 0.5)
p <- p + ggplot2::facet_wrap("method", nrow=2)
p <- p + paletteer::scale_fill_paletteer_d(
  palette = "ggthemes::Tableau_10",
  drop = FALSE)
p <- p + ggplot2::scale_y_sqrt(name = latex2exp::TeX("$| \\hat{\\lambda}^r - \\hat{\\lambda}_{0}|$"))
p <- p + ggplot2::xlab("weighting method")
p <- p + ggplot2::theme(
  axis.text.x = ggplot2::element_text(
    angle=60.0,
    hjust=1.0))

@@

\midrule
non-robust               & ---  & ---       & ---  & ---  & ---       & ---  & 783 \\
$p^{*}$ (step)           & 0.80 & $(0, 1]$  & 0.84 & ---  & ---       & ---  & 587 \\
$p^{*}$ (triangle)       & 0.80 & $(0, 1]$  & 0.78 & 0.95 & $(0, 1]$  & 0.99 & 587 \\
$p^{*}$ (tapered cosine) & 0.80 & $(0, 1]$  & 0.76 & 0.95 & $(0, 1]$  & 0.98 & 583 \\
$z$ (step)               & 1.28 & $(0, 10]$ & 1.09 & ---  & ---       & ---  & 1135 \\
$z$ (triangle)           & 1.28 & $(0, 10]$ & 0.26 & 1.96 & $(0, 10]$ & 5.47 & 1813 \\
$z$ (tapered cosine)     & 1.28 & $(0, 10]$ & 0.03 & 1.96 & $(0, 10]$ & 6.91 & 1844 \\
$r$ (step)               & 0.50 & $(0, 10]$ & 1.56 & ---  & ---       & ---  & 1837 \\
$r$ (triangle)           & 0.50 & $(0, 10]$ & 1.49 & 1.00 & $(0, 10]$ & 1.51 & 1877 \\
$r$ (tapered cosine)     & 0.50 & $(0, 10]$ & 1.32 & 1.00 & $(0, 10]$ & 1.65 & 1834 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```

@@

\midrule
non-robust               & ---  & ---       & ---   & ---  & ---       & ---   & 225 \\
$p^{*}$ (step)           & 0.80 & $(0, 1]$  & 0.92  & ---  & ---       & ---   & 125 \\
$p^{*}$ (triangle)       & 0.80 & $(0, 1]$  & 0.88  & 0.95 & $(0, 1]$  & 0.93  & 126 \\
$p^{*}$ (tapered cosine) & 0.80 & $(0, 1]$  & 0.89  & 0.95 & $(0, 1]$  & 0.93  & 124 \\
$z$ (step)               & 1.28 & $(0, 10]$ & 1.04  & ---  & ---       & ---   & 333 \\
$z$ (triangle)           & 1.28 & $(0, 10]$ & 0.12  & 1.96 & $(0, 10]$ & 9.91  & 620 \\
$z$ (tapered cosine)     & 1.28 & $(0, 10]$ & 0.23  & 1.96 & $(0, 10]$ & 5.63  & 716 \\
$r$ (step)               & 0.50 & $(0, 10]$ & 10.00 & ---  & ---       & ---   & 485 \\
$r$ (triangle)           & 0.50 & $(0, 10]$ & 9.96  & 1.00 & $(0, 10]$ & 10.00 & 483 \\
$r$ (tapered cosine)     & 0.50 & $(0, 10]$ & 9.97  & 1.00 & $(0, 10]$ & 10.00 & 484 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```

Table **INSERT** shows weighting parameters and corresponding loss, after optimisation. Loss without any mitigation of outliers was $225$. Only weighting based on empirical probabilities ($p^(*)$) achieved better loss under optimised weighting parameters. For these data, all three window functions achieved a similar loss. Weighting based on the z-score of the transformed feature, or on their residual error, performed worse than non-robust shift-sensitive power transformation. Thus weighting of outliers should not be based on either of these two options. The likely reason for their poor performance is that they allow for a direct interaction with estimation of the $\lambda$ parameter. Because they both use transformed feature values, weights actually vary during the optimisation process for maximising the weighted log-likelihood function. This complicates the parameter search space, and leads to different parameter estimates. Weighting based on empirical probabilities does not suffer from this issue, because all weights are fixed during optimising the weighted log-likelihood function.
For weighting based on the z-score of the transformed feature, or on their residual error, we can see two extreme strategies emerging. Firstly, the strategy to down-weight most instances ($r$ (step), and to a lesser degree $z$ (step)) and thereby force normality on the most central parts of the distribution. Second is an attempt to assign full-weight to almost all instances, with large values for $k_1$ and $k_2$, which leads to behaviour similar to non-robust power transformation.

>>>>

.plot_weighting_functions(plot_theme = plot_theme)
```
Other criteria aside from MLE can be used for finding robust transformation parameters. We define robust criteria in Appendix A **insert**.

@@

The result is shown in Figure **INSERT**. For each distribution, transformation parameter $\hat{\lambda}$ varies with $d$ when estimated using the original definitions. In contrast, estimation of $\lambda$ using shift-sensitive power transformations was invariant to $d$.

@@

The optimal weighting function parameters for weighted MLE are shown in Tables **insert** and **insert** for robust, shift-sensitive Box-Cox and Yeo-Johnson transformations, respectively.

@@

\midrule
non-robust               & ---  & ---       & ---  & ---  & ---       & ---  & 773 \\
$p^{*}$ (step)           & 0.80 & $(0, 1]$  & 0.84 & ---  & ---       & ---  & 598 \\
$p^{*}$ (triangle)       & 0.80 & $(0, 1]$  & 0.79 & 0.95 & $(0, 1]$  & 0.98 & 594 \\
$p^{*}$ (tapered cosine) & 0.80 & $(0, 1]$  & 0.77 & 0.95 & $(0, 1]$  & 1.00 & 580 \\
$z$ (step)               & 1.28 & $(0, 10]$ & 1.21 & ---  & ---       & ---  & 1214 \\
$z$ (triangle)           & 1.28 & $(0, 10]$ & 1.05 & 1.96 & $(0, 10]$ & 4.47 & 1052 \\
$z$ (tapered cosine)     & 1.28 & $(0, 10]$ & 1.09 & 1.96 & $(0, 10]$ & 5.30 & 1086 \\
$r$ (step)               & 0.50 & $(0, 10]$ & 1.36 & ---  & ---       & ---  & 1483 \\
$r$ (triangle)           & 0.50 & $(0, 10]$ & 1.43 & 1.00 & $(0, 10]$ & 1.45 & 1498 \\
$r$ (tapered cosine)     & 0.50 & $(0, 10]$ & 1.20 & 1.00 & $(0, 10]$ & 1.48 & 1467 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```

@@

\midrule
non-robust               & ---  & ---       & ---  & ---  & ---       & ---  & 305 \\
$p^{*}$ (step)           & 0.80 & $(0, 1]$  & 0.87 & ---  & ---       & ---  & 136 \\
$p^{*}$ (triangle)       & 0.80 & $(0, 1]$  & 0.83 & 0.95 & $(0, 1]$  & 0.88 & 133 \\
$p^{*}$ (tapered cosine) & 0.80 & $(0, 1]$  & 0.81 & 0.95 & $(0, 1]$  & 0.88 & 133 \\
$z$ (step)               & 1.28 & $(0, 10]$ & 1.07 & ---  & ---       & ---  & 333 \\
$z$ (triangle)           & 1.28 & $(0, 10]$ & 1.00 & 1.96 & $(0, 10]$ & 4.57 & 227 \\
$z$ (tapered cosine)     & 1.28 & $(0, 10]$ & 0.55 & 1.96 & $(0, 10]$ & 5.16 & 236 \\
$r$ (step)               & 0.50 & $(0, 10]$ & 1.57 & ---  & ---       & ---  & 631 \\
$r$ (triangle)           & 0.50 & $(0, 10]$ & 1.28 & 1.00 & $(0, 10]$ & 1.81 & 562 \\
$r$ (tapered cosine)     & 0.50 & $(0, 10]$ & 1.32 & 1.00 & $(0, 10]$ & 1.75 & 567 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```

Figure **insert** shows the distribution of errors $\left| \hat{\lambda}^r - \hat{\lambda}_0 \right|$ for non-robust transformations, as well as robust transformations using the optimal weighting function parameters. In the presence of outliers on either side of the distribution, the Yeo-Johnson transformation yields smaller errors than the Box-Cox transformation. For both transformations, weighting based on empirical probabilities yields the most consistent $\lambda$ parameter estimates in the presence of outliers. Weighting based on residual errors tended to perform least well, and were generally similar to (Yeo-Johnson) or worse than (Box-Cox) non-robust transformations.

```{r optimised_weighting_function_parameters, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Robustness of power transformations after optimising weighting function parameters. ",
  "The distribution of errors, i.e. the difference between robustly fitted $\\hat{\\lambda}^r$ ",
  "and expected $\\hat{\\lambda}_{0}$ found in the absence of outliers, is shown for 1000 randomly ",
  "parametrised asymmetric generalised normal distributions with up to 10% outliers on either side ",
  "of the distribution. The top and bottom panel show errors for the shift-sensitive Box-Cox transformation, ",
  "and shift-sensitive Yeo-Johnson transformation, respectively. ",
  "In each panel, the error distribution for the non-robust transformation is shown to the left ",
  "for comparison with different weighting functions. ",
  "Note that a square root transformation was used along the $y$-axis. Median errors are shown above each distribution."
)

.plot_optimised_weighting_function_parameters(
  plot_theme = plot_theme,
  manuscript_dir = manuscript_dir)
```
